An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: What Role for the Human in Higher Education?

### 2025-01-24 paper/paper.qmd
Revisions for possible submission to Critical AI

The use of Generative Artificial Intelligence (AI) in higher education has been a sensational topic since the launch of ChatGPT and similar tools. AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students. This presentation of AI is misleading, and there is a need to examine the claims about what the technology can do and what it actually does in the context of teaching and learning. In this paper I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma). However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.

@@

On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022]. Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].], and no surprise, given how the technology was hailed as nothing short of revolutionary.

@@

AI was quickly painted as an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. Institutions of higher education were compelled to get in on the act, forming high-level advisory committees and drafting guidelines to guide the university community leverage AI in principled ways [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities. And yet not every single one commands a high-level committee to advice on its use. Just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, examine what it actually does.

@@

This essay aims to provide some clarification around 1) the things that GenAIs claim to do in education; and 2) what the technology actually does; within 3) the context of a principled organization. I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.

@@

This overarching goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].

Unprincipled organizations do not need to question what the adoption of a technology does to and for whom--as long as it advances whatever its goal is at any particular moment. But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

@@

Neural networks and machine learning are nothing new, and they have their uses. But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI. This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]. There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.

@@

I hope to have conveyed the sense that AI as marketed by the likes of OpenAI and Microsoft, cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability. This is well-known: most discussions of AI in education acknowledge that their use is rife with problems. For instance, @mollick_instructors_2024 note issues with copyright, exploitation of low-wage workers, reproduction of biases, the risk of automation bias, and privacy loss. These are issues that "educators may want to consider before deciding whether to use these systems". This disclaimer, consuming fully one third of a page in a 76-page document, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual educators working under desperately under-resourced conditions, or simply willing to win a race to the bottom. And yet, I have little illusion that these arguments will prove persuasive.

I would still appeal, though, to the university's self-regard and sense of self-preservation.

We would do our students an unforgivable disservice by offering them a skill that reversed them to mediocrity and left them at the mercy of predatory corporations. "This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education. But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far less expensive alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").

Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors, a university's main source of authority is its reputation as a place that preserves and expands knowledge in a principled way. Give the principle away, what is left to sustain universities as institutions that add value to society?

>>>>

The use of Generative Artificial Intelligence (AI) in higher education has been a sensational topic since the launch of ChatGPT and similar tools. AI has been presented as an everything tool that can do a fabulous array of tasks, from creating content to grading to providing feedback to students. This presentation of AI is misleading, and there is a need to examine the claims about what the technology can do and what it actually does in the context of teaching and learning. In this essay I scan the academic and gray literature, along with media reports regarding AI, to show how discussions about AI hinge on making a large number of unverified claims (Gish Gallop) while presenting the technology as inevitable (a False Dilemma). However, once we examine what AI does, and particularly what it does for whom and to whom, a picture emerges of a technology that is intrinsically inimical to the principles that most institutions of higher education claim to hold dear, including the best interest of their students.

@@

On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022]. Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].].

@@

AI was quickly painted as nothing short of revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. Institutions of higher education were compelled to get in on the act, forming high-level advisory committees and drafting guidelines to guide the university community leverage AI in principled ways [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities. And yet not every single one commands a high-level university committee to advice on their use. Just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, examine what it actually does.

@@

This essay aims to provide some clarification around 1) claims about the potential of GenAIs in education; and 2) what GenAIs actually do--within 3) the context of a principled organization. I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.

@@

This goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].

Unprincipled organizations do not need to question their choices of technology--as long as the technology advances whatever its goal is at any particular moment. But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

@@

Neural networks and machine learning are nothing new, and they have their uses. But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI. This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools willing to believe in magic mirrors [@burneko_if_2024]. There are now calls to press pause on the hype [@angwin_press_2024] but these voices are barely audible above the thundering echos inside the AI hype chamber.

@@

AI, as marketed by the likes of OpenAI, Googla, and Microsoft, cannot be used in a way that is consistent with values like integrity, respect for human rights, and environmental sustainability. This is well-known: most discussions of AI in education acknowledge that their use is rife with problems. For instance, @mollick_instructors_2024 note issues with copyright, exploitation of low-wage workers, reproduction of biases, the risk of automation bias, and privacy loss. These are issues that "educators may want to consider before deciding whether to use these systems". This disclaimer highlights the important role of institutions to provide the moral and ethical clarity that may elude individual educators working under desperately under-resourced conditions, or simply willing to win a race to the bottom. And yet, I have little illusion that these arguments will prove persuasive.

I would still appeal, though, to the university's self-regard and sense of self-preservation. 
We would do our students an unforgivable disservice by offering them a skill that reversed them to mediocrity and left them at the mercy of predatory corporations. "This university uses ChatGPT to teach" is simply a way of saying that we are too cheap to offer a human-centered education. But in truth, a university is too expensive an institution to be attractive under such conditions, when there are far less expensive alternatives for an AI-based education ("for only $9.99 proof yourself against layoffs").

Going down this path would be a regrettable mistake at a time when universities are under siege from unsympathetic or actively hostile actors. A university’s main source of authority is its reputation as a place that preserves and expands knowledge in a principled way. If we give this principle away, what would be left to sustain universities as institutions that add value to society?
